The effect of climate forcing on numerical simulations of the Cordilleran Ice Sheet at the Last Glacial Maximum

### 2013-11-12 sections/abstract.tex
More small corrections from Irina

\begin{abstract}
We present an ensemble of numerical simulations of the Cordilleran Ice Sheet during the Last Glacial Maximum performed with the Parallel Ice Sheet Model (PISM), applying temperature offsets to the present-day climatologies from five different datasets. Monthly mean surface air temperature and precipitation from WorldClim, the NCEP/NCAR reanalysis, the ERA-Interim reanalysis, the Climate Forecast System Reanalysis and the North American Regional Reanalysis are used to compute surface mass balance in a positive degree-day model. Modelled ice sheet outlines and volumes appear highly sensitive to the choice of climate forcing. For three of the four reanalysis dataset used, differences in precipitation is the major source for discrepancies between model results. We assess model performance against a geomorphological reconstruction of the ice margin at the Last Glacial Maximum, and suggest that part of the mismatch is due to unresolved orographic precipitation effects caused by the coarse resolution of reanalysis data. The best match between model output and the reconstructed ice margin is obtained using the high-resolution North American Regional Reanalysis, which we retain for future simulations of the Cordilleran Ice Sheet.
\end{abstract}

>>>>

\begin{abstract}
We present an ensemble of numerical simulations of the Cordilleran Ice Sheet during the Last Glacial Maximum performed with the Parallel Ice Sheet Model (PISM), applying temperature offsets to the present-day climatologies from five different datasets. Monthly mean surface air temperature and precipitation from WorldClim, the NCEP/NCAR reanalysis, the ERA-Interim reanalysis, the Climate Forecast System Reanalysis and the North American Regional Reanalysis are used to compute surface mass balance in a positive degree-day model. Modelled ice sheet outlines and volumes appear highly sensitive to the choice of climate forcing. For three of the four reanalysis datasets used, differences in precipitation are the major source for discrepancies between model results. We assess model performance against a geomorphological reconstruction of the ice margin at the Last Glacial Maximum, and suggest that part of the mismatch is due to unresolved orographic precipitation effects caused by the coarse resolution of reanalysis data. The best match between model output and the reconstructed ice margin is obtained using the high-resolution North American Regional Reanalysis, which we retain for simulations of the Cordilleran Ice Sheet in the future.
\end{abstract}

### 2013-11-12 sections/climate.tex
More small corrections from Irina

WorldClim is a high resolution global climate dataset built from meteorological observations \citep{data:worldclim}. It was built by applying a spatial interpolation between a large set of measurements taken at terrestrial weather stations worldwide, using SRTM \citep{data:srtm} and GTOPO30 elevation data \citep{data:gtopo30}. This dataset provides a resolution much higher than attained by circulation models, however its reliability is variable to the highly inhomogeneous distribution of measurements \citep{data:worldclim}.

Within our modelling domain, the spatial distributions of summer (JJA) and winter (DJF) air surface temperatures in WorldClim generally show a decrease of temperature from south to north (JJA) and from south-west to north-east (DJF), such that regions further inland experience colder winters (Fig.~\ref{fig:temp}). It should be noted that temperature gradients are reversed (colder inland than at the coast) and much stronger in winter than in summer. Temperatures rises well above zero during the summer months over the entire modelling domain, except for the highest mountain peaks, and even regions where mean annual temperatures are well below freezing point do experience warm summers. In other words, there is a strong contrast in seasonality between coastal temperate-climate and inland continental-climate regions.

The spatial distribution of JJA and DJF mean precipitation rates in WorldClim reveal a strong precipitation decline beyond the coastal regions (Fig.~\ref{fig:prec}), primarily as a result of the continuous orographic barrier formed by the Boundary Ranges, the Coast Mountains and the Cascades. A less pronounced precipitation peak is also visible over the interior ranges. From an ice sheet mass balance point of view, these heterogeneities become even stronger due to the difference in timing of the precipitation peak through the year. Whereas coastal regions experience most precipitation as snow during the accumulation season (DJF), inland regions experience dry winters and most of the annual precipitation consequently falls as rain during the ablation season (JJA).

@@

The spatial distributions of JJA air surface temperatures and DJF precipitation rates from the four reanalyses climatologies are shown in Figures~\ref{fig:temp} and~\ref{fig:prec}. Summer temperature and winter precipitation are most relevant to the glacier model as they respectively drive summer melt and winter accumulation. Because reanalyses include observational information through data assimilation, they are closer to observations than pure GCMs in densely monitored regions, such as the southern part of our modelling domain \citep{bengtsson-etal-2007}. On the other hand, the atmospheric physics embedded in the models provide more realistic output than obtained by spatial interpolation in sparsely monitored regions, such as the northern part of our modelling domain.

>>>>

WorldClim is a high resolution global climate dataset built from meteorological observations \citep{data:worldclim}. It was built by applying a spatial interpolation between a large set of measurements taken at terrestrial weather stations worldwide, using SRTM \citep{data:srtm} and GTOPO30 elevation data \citep{data:gtopo30}. This dataset provides a resolution much higher than attained by circulation models, however its reliability is variable due to the highly inhomogeneous distribution of measurements \citep{data:worldclim}.

Within our modelling domain, the spatial distributions of summer (JJA) and winter (DJF) air surface temperatures in WorldClim generally show a decrease of temperature from south to north (JJA) and from south-west to north-east (DJF), such that regions further inland experience colder winters (Fig.~\ref{fig:temp}). It should be noted that temperature gradients are much stronger in winter than in summer. Temperature rises well above zero during the summer months over the entire modelling domain, except for the highest mountain peaks, and even regions where mean annual temperatures are well below freezing point do experience warm summers. In other words, there is a strong contrast in seasonality between coastal temperate-climate and inland continental-climate regions.

The spatial distribution of JJA and DJF mean precipitation rates in WorldClim reveals a strong precipitation decline beyond the coastal regions (Fig.~\ref{fig:prec}), primarily as a result of the continuous orographic barrier formed by the Boundary Ranges, the Coast Mountains and the Cascades. A less pronounced precipitation peak is also visible over the interior ranges. From an ice sheet mass balance point of view, these heterogeneities become even stronger due to the difference in timing of the precipitation peak through the year. Whereas coastal regions experience most precipitation as snow during the accumulation season (DJF), inland regions experience dry winters and most of the annual precipitation consequently falls as rain during the ablation season (JJA).

@@

The spatial distributions of JJA air surface temperatures and DJF precipitation rates from the four reanalysis climatologies are shown in Figures~\ref{fig:temp} and~\ref{fig:prec}. Summer temperature and winter precipitation are most relevant to the glacier model as they respectively drive summer melt and winter accumulation. Because reanalyses include observational information through data assimilation, they are closer to observations than pure GCMs in densely monitored regions, such as the southern part of our modelling domain \citep{bengtsson-etal-2007}. On the other hand, the atmospheric physics embedded in the models provide more realistic output than obtained by spatial interpolation in sparsely monitored regions, such as the northern part of our modelling domain.

### 2013-11-12 sections/concl.tex
More small corrections from Irina

For the Cordilleran Ice Sheet, we achieve the best fit to the mapped LGM margin by \citet{dyke-2004} using climate forcing from the high-resolution interpolated observational data WorldClim \citep{data:worldclim} and the North American Regional Reanalysis \citep[NARR;][]{data:narr}. The latter dataset is preferable in our case due to the lack of WorldClim data offshore. Climate forcing from CFSR and the NCEP/NCAR reanalysis produces largely oversized ice sheets due to too high precipitation rates, and in the second case, too low surface air temperature. The ERA-Interim data used in this study produces more reasonable results, but it may be too coarse to accurately resolve the spatial distribution of orographic precipitation associated with the rugged topography of the northern American Cordillera, resulting in a misplaced ice cover in regard to geomorphological data. Therefore, we retain NARR data for forcing future simulations of the Cordilleran Ice Sheet.

>>>>

For the Cordilleran Ice Sheet, we achieve the best fit to the mapped LGM margin by \citet{dyke-2004} using climate forcing from the high-resolution interpolated observational data WorldClim \citep{data:worldclim} and the North American Regional Reanalysis \citep[NARR;][]{data:narr}. The latter dataset is preferable in our case due to the lack of WorldClim data offshore. Climate forcing from CFSR and the NCEP/NCAR reanalysis produce largely oversized ice sheets due to too high precipitation rates, and in the second case, too low surface air temperature. The ERA-Interim data used in this study produces more reasonable results, but it may be too coarse to accurately resolve the spatial distribution of orographic precipitation associated with the rugged topography of the northern American Cordillera, resulting in a misplaced ice cover in regard to geomorphological data. Therefore, we retain NARR data for forcing simulations of the Cordilleran Ice Sheet in the future.

### 2013-11-12 sections/discuss.tex
More small corrections from Irina

Although this selection result in similarly sized ice sheets, noticeable differences in shape exist. The ERA-Interim simulation produces a ore northerly-centred ice sheet with too much ice in the North and too little in the South. This may partly reflect unresolved orographic precipitation effects (Fig.~\ref{fig:oroprecip}) due to coarser GCM resolution (Table~\ref{tab:reanalyses}).

@@

Firstly, the modelled eastern margin of the ice sheet extends further east than where one would expect junction between the Cordilleran and the Laurentide ice sheets. This can be seen for instance in the south-eastern part of our modelling domain, where the modelled ice sheets cover an ice-free corridor of the LGM reconstruction (Fig.~\ref{fig:best}). However the Laurentide Ice Sheet, absent from our simulations, likely formed a buttress against the smaller Cordilleran Ice Sheet and stopped its advance onto the Canadian Prairies. In addition, potential effects of the growing ice-sheet on regional climate are not included in our model. The Cordilleran ice sheet initiated from the junction of mountain ice caps over the major reliefs \citep{clague-1989}. In our simulations, a continuous ice cover quickly forms over the western ranges, where precipitation rates are higher than in the rest of the domain. This continuous ice cover may have enhanced the topographical barrier already formed by the western ranges, resulting in less precipitation and warmer air in the interior. This process of ``self-inhibiting growth'' was demonstrated in a similar setting over Greenland, where it may have been limiting the advance of the ice sheet during its build-up \citep{langen-etal-2012}.

@@

Considering the lack of WorldClim data offshore, we consider the NARR climate forcing as best suited for simulations of the Cordilleran Ice Sheet among those tested, and under the assumptions of the present study.

@@

To test the effect of a shorter englaciation period on model results, we use simulations run under NARR climate forcing with temperatures offsets of 7 to 11\,\unit{\degree C}, and compare modelled ice sheet extent when glaciated area reaches the approximate size of the LGM Cordilleran Ice Sheet of $2\,\times10^6\,\unit{km^2}$.

@@

Furthermore, some simplifications were made in the surface mass balance model. Our PDD model does not include refreezing. Refreezing of melted snow and ice and retention in the snow pack, however, can greatly alter mass balance at the surface of an ice sheet \citep{janssens-huybrechts-2000}. Additionally, we simulate temperature variability by using a constant, uniform value of temperature standard deviation. This approach implies large biases of surface mass-balance \citep{charbit-etal-2013,rau-rogozhina-2013,seguinot-inpress}, particularly over regions with such various climates as the Northern American Cordillera.

>>>>

Although this selection results in similarly sized ice sheets, noticeable differences in shape exist. The ERA-Interim simulation produces a more northerly-centred ice sheet with too much ice in the North and too little in the South. This may partly reflect unresolved orographic precipitation effects (Fig.~\ref{fig:oroprecip}) due to coarser GCM resolution (Table~\ref{tab:reanalyses}).

@@

Firstly, the modelled eastern margin of the ice sheet extends further east than where one would expect junction between the Cordilleran and the Laurentide ice sheets. This can be seen for instance in the south-eastern part of our modelling domain, where the modelled ice sheets cover an ice-free corridor of the LGM reconstruction (Fig.~\ref{fig:best}). However the Laurentide Ice Sheet, absent from our simulations, likely formed a buttress against the smaller Cordilleran Ice Sheet and stopped its advance onto the Canadian Prairies. In addition, potential effects of the growing ice-sheet on atmospheric circulation and precipitation changes are not included in our model. The Cordilleran ice sheet initiated from the junction of mountain ice caps over the major reliefs \citep{clague-1989}. In our simulations, a continuous ice cover quickly forms over the western ranges, where precipitation rates are higher than in the rest of the domain. This continuous ice cover may have enhanced the topographical barrier already formed by the western ranges, resulting in less precipitation and warmer air in the interior. This process of ``self-inhibiting growth'' was demonstrated in a similar setting over Greenland, where it may have been limiting the advance of the ice sheet during its build-up \citep{langen-etal-2012}.

@@

Given the lack of WorldClim data offshore, we consider the NARR climate forcing as best suited for simulations of the Cordilleran Ice Sheet among those tested, and under the assumptions of the present study.

@@

To test the effect of a shorter englaciation period on model results, we use simulations run under NARR climate forcing with temperature offsets of 7 to 11\,\unit{\degree C}, and compare modelled ice sheet extent when glaciated area reaches the approximate size of the LGM Cordilleran Ice Sheet of $2\,\times10^6\,\unit{km^2}$.

@@

Furthermore, some simplifications were made in the surface mass balance model. Our PDD model does not include refreezing. Refreezing of melted snow and ice and retention in the snow pack, however, can greatly alter mass balance at the surface of an ice sheet \citep{janssens-huybrechts-2000}. Additionally, we simulate temperature variability by using a constant, uniform value of daily temperature standard deviation. This approach implies large biases of surface mass-balance \citep{charbit-etal-2013,rau-rogozhina-2013,seguinot-inpress}, particularly over regions with such various climates as the Northern American Cordillera.

### 2013-11-12 sections/intro.tex
More small corrections from Irina

Here, we propose to address some of the uncertainties concerning climate forcing of numerical glacier models by evaluating the responses of a numerical model, in terms of glacier extent, to inputs fromo several climate datasets. This is not entirely a new approach as a few studies of this kind are presently available. \citet{quiquet-etal-2012}, for example, assessed the sensitivity of a Greenland ice sheet model to various atmospheric forcing, including a regional parametrization by \citet{fausto-etal-2009}, output from several GCMs and an atmospheric reanalysis. \citet{rodgers-etal-2004} and \citet{charbit-etal-2007} tested the sensitivity of a model of the Northern Hemisphere ice sheet model to climate forcing from different PMIP LGM simulations. Although using different set-ups, all three studies demonstrate the very large sensitivity of ice sheet model to the choice of climate forcing data used. In our study, to limit the degrees of freedom in our model and obtain results independent of palaeo-ice sheet reconstructions such as ICE-4G, we use a simple temperature offset similar to the approaches by \citet{greve-etal-1999} and \citet{bintanja-etal-2005} and assess ice sheet model sensitivity to the choice of present-day climate data. Hence, rather than using GCM output, we force our model with climate reanalysis data, which builds on observational information through data assimilation \citep{bengtsson-etal-2007}. Furthermore, we focus our study regionally on the former Cordilleran Ice Sheet in western Northern America.

The Cordilleran Ice Sheet (Fig.~\ref{fig:locmap}) covered an area that presently experiences strong regional variations in climate. From a numerical modelling perspective, it is one of the least studied palaeo-ice sheets of the Northern Hemisphere, despite the fact that significant stratigraphical and geomorphological data is available to constrain its extent \citep{jackson-clague-1991,dukrodkin-1999,kaufman-manley-2004,kleman-etal-2010,margold-etal-2011}. Although the Cordilleran Ice Sheet was not the primary target of these simulations, it has previously been modelled as part of efforts to simulate ice sheets in North America \citep{marshall-clarke-1999,calov-etal-2002,tarasov-peltier-1997,tarasov-peltier-2004,gregoire-etal-2012}, the Northern Hemisphere \citep{huybrechts-tsiobbel-1996,greve-etal-1999,charbit-etal-2002,charbit-etal-2007,charbit-etal-2013,johnson-fastook-2002,rodgers-etal-2004,bintanja-etal-2005,zweck-huybrechts-2005,abeouchi-etal-2007} and world-wide \citep{yoshimori-etal-2001}. While these studies generally reproduce the magnitude of North American glaciation at LGM reasonably well, there exists a tendency in the simulations that are independent of ice sheet reconstructions such as the ICE-4G to predict excessive ice cover in parts of northern Yukon Territory and interior Alaska that have remained ice-free throughout the Pleistocene \citep{dukrodkin-1999,kaufman-manley-2004}.

>>>>

Here, we propose to address some of the uncertainties concerning climate forcing of numerical glacier models by evaluating the responses of a numerical model, in terms of glacier extent, to inputs from several climate datasets. This is not entirely a new approach as a few studies of this kind are presently available. \citet{quiquet-etal-2012}, for example, assessed the sensitivity of a Greenland ice sheet model to various atmospheric forcing, including a regional parametrization by \citet{fausto-etal-2009}, output from several GCMs and an atmospheric reanalysis. \citet{rodgers-etal-2004} and \citet{charbit-etal-2007} tested the sensitivity of a model of the Northern Hemisphere ice sheet model to climate forcing from different PMIP LGM simulations. Although using different set-ups, all three studies demonstrate the very large sensitivity of ice sheet model to the choice of climate forcing data used. In our study, to limit the degrees of freedom in our model and obtain results independent of palaeo-ice sheet reconstructions such as ICE-4G, we use a simple temperature offset similar to the approaches by \citet{greve-etal-1999} and \citet{bintanja-etal-2005} and assess ice sheet model sensitivity to the choice of present-day climate data. Hence, rather than using GCM output, we force our model with climate reanalysis data, which builds on observational information through data assimilation \citep{bengtsson-etal-2007}. Furthermore, we focus our study regionally on the former Cordilleran Ice Sheet in western Northern America.

The Cordilleran Ice Sheet (Fig.~\ref{fig:locmap}) covered an area that presently experiences strong regional variations in climate. From a numerical modelling perspective, it is one of the least studied palaeo-ice sheets of the Northern Hemisphere, despite the fact that significant stratigraphical and geomorphological data is available to constrain its extent \citep{jackson-clague-1991,dukrodkin-1999,kaufman-manley-2004,kleman-etal-2010,margold-etal-2011}. The Cordilleran Ice Sheet has previously been modelled as part of efforts to simulate ice sheets in North America \citep{marshall-clarke-1999,calov-etal-2002,tarasov-peltier-1997,tarasov-peltier-2004,gregoire-etal-2012}, the Northern Hemisphere \citep{huybrechts-tsiobbel-1996,greve-etal-1999,charbit-etal-2002,charbit-etal-2007,charbit-etal-2013,johnson-fastook-2002,rodgers-etal-2004,bintanja-etal-2005,zweck-huybrechts-2005,abeouchi-etal-2007} and world-wide \citep{yoshimori-etal-2001}. While these studies generally reproduce the magnitude of North American glaciation at LGM reasonably well, there exists a tendency in the simulations that are independent of ice sheet reconstructions such as the ICE-4G to predict excessive ice cover in parts of northern Yukon Territory and interior Alaska that have remained ice-free throughout the Pleistocene \citep{dukrodkin-1999,kaufman-manley-2004}.

### 2013-11-12 sections/model.tex
Fix parentheses in references
More small corrections from Irina

Given basal topography, sea level, geothermal heat flux and climate forcing, the model computes ice extent and thickness, its thermal and dynamic state, and the associated lithospheric response. Our modelling domain encompasses the entire area covered by the Cordilleran Ice Sheet at LGM including independent ice growing on the western Alaskan ranges and the Brooks Range and intermediated expanses of ice-free terrain in northern Yukon Territory and interior Alaska (Fig.~\ref{fig:locmap}).

@@

The central part of an ice sheet model consist of the computation of flow velocity which itself depends on temperature. PISM is a shallow model, which implies that the balance of stresses is approximated based on their predominant components. On the other hand, the model is polythermal: it accounts for differences in temperature and softness within the ice column.

@@

A pseudo-plastic sliding law \citep{aschwanden-etal-2013} (supplement) relates the bed-parallel shear stress and the sliding velocity. The yield stress is modelled using the Mohr-Coulomb criterion. The till friction angle $\phi$ varies from 10 to 30\degree\ and is a function of current bed elevation, with lowest values occurring at low elevations,

>>>>

Given basal topography, sea level, geothermal heat flux and climate forcing, the model computes ice extent and thickness, its thermal and dynamic state, and the associated lithospheric response. Our modelling domain encompasses the entire area covered by the Cordilleran Ice Sheet at LGM including independent ice growing on the western Alaskan ranges and the Brooks Range and intermediate expanses of ice-free terrain in northern Yukon Territory and interior Alaska (Fig.~\ref{fig:locmap}).

@@

The central part of an ice sheet model consists of the computation of flow velocity which itself depends on temperature. PISM is a shallow model, which implies that the balance of stresses is approximated based on their predominant components. On the other hand, the model is polythermal: it accounts for differences in temperature and softness within the ice column.

@@

A pseudo-plastic sliding law \citep[supplement]{aschwanden-etal-2013} relates the bed-parallel shear stress and the sliding velocity. The yield stress is modelled using the Mohr-Coulomb criterion. The till friction angle $\phi$ varies from 10 to 30\degree\ and is a function of current bed elevation, with lowest values occurring at low elevations,

### 2013-11-12 sections/results.tex
More small corrections from Irina

To contrast the effect of the six different input climatologies, we plot simulations using a 5\,\unit{\degree C} temperature offset (Fig.~\ref{fig:cool05}). The magnitude of glaciation differs dramatically from one climate forcing to another. Whereas the NCEP/NCAR and CFSR climatologies produce a large ice sheet that covers most of the model domain (Fig.~\ref{fig:cool05}, lower panels), WorldClim, ERA-Interim and NARR climatologies lead to more restrictive ice cover (Fig.~\ref{fig:cool05}, upper panels). The WorldClim run only resulted in the growth of mountain ice caps bounded to the highest topography.

>>>>

To contrast the effect of the six different input climatologies, we plot simulations using a 5\,\unit{\degree C} temperature offset (Fig.~\ref{fig:cool05}). The magnitude of glaciation differs dramatically from one climate forcing to another. Whereas the NCEP/NCAR and CFSR climatologies produce a large ice sheet that covers most of the model domain (Fig.~\ref{fig:cool05}, lower panels), WorldClim, ERA-Interim and NARR climatologies lead to a more restrictive ice cover (Fig.~\ref{fig:cool05}, upper panels). The WorldClim run only resulted in the growth of mountain ice caps bounded to the highest topography.
